---
title: "2024 US Men's Gymnastics Olympic Team Selection"
author: "Dante Yasui"
date: "2024/06/29"
format:
  html:
    df-print: kable
    toc: true
---

# TODO

- [ ] make events print in olympic order, not alphabetic
- [ ] calculate each team members contribution by picking best team 
when they are dropped from results dataset

---

# Data

I copied the data from <https://myusagym.com>
and manually cleaned it up into `.csv` format.

```{r}
#| message: FALSE
#| lst-label: lst-data
#| lst-cap: combining data from nationals and trials
library(tidyverse)
library(furrr)
library(future)
# read all results across usa champs and olympic trials
nc1_df <- read_csv("./natnls-d1.csv")
nc2_df <- read_csv("./natnls-d2.csv")
ot1_df <- read_csv("./trials-day1.csv")
ot2_df <- read_csv("./trials-day2.csv")
# merge all results into one dataframe
results_df <- bind_rows(list(nc1_df, nc2_df, ot1_df, ot2_df))
```

To visualize the data overall, 
I grouped scores by event from all USA Championships and Olympic Trials 
routines and plotted the spread of observed scores.

```{r}
#| label: fig-eventdist
#| fig-cap: national distributions of scores on each event
#| lst-label: lst-eventdist
#| lst-cap: plot of event distributions
library(ggthemes)
eventsdist <- results_df %>% select(!AA) %>%
  pivot_longer(
    cols = c(FX, PH, SR, VT, PB, HB),
    names_to = "Event",
    values_to = "Score",
  ) %>% filter(Score > 1) %>%
  ggplot(aes(x=Score)) +
  geom_density() +
  xlim(11, 16) +
  facet_wrap(~Event, ncol = 2) +
  theme_pander()
eventsdist
```

Not surprisingly, out of all six event Pommel has the largest variance in
scores and the lowest average.
P-Bars and Vault are high-scoring events on average but Vault 
(at least for these competitions) is more consistently grouped around 14
while the P-Bars distribution seem to have a large right tail 
with a lot of potential for 15+ scores.

## Averages across usa championships and olympic trials

```{r}
#| label: tbl-averagescores
#| tbl-cap: "Individuals' average scores across nationals and trials"
# average individual gymnasts' scores across four days
indv.averages_df <- results_df %>% 
  filter(NAME %in% ot1_df$NAME) %>%  # who qualified to olympic trials 
  group_by(NAME) %>% 
  summarise_at(vars(FX:AA), mean, na.rm=TRUE) %>% 
  arrange(desc(AA))
indv.averages_df
```

---

## Predicting a Team's performance
```{r}
#| lst-label: lst-functions
#| lst-cap: functions for calculating team performance

# function to pick top three scores on each event from a team
team_top_scores <- function(names, scores = indv.averages_df, perevent = 3) {
  scores %>% 
    filter(NAME %in% names) %>%
    select(-AA) %>% 
    pivot_longer(
      -NAME,
      names_to = "Event",
      values_to = "Score",
    ) %>% 
    # take top 3 on each event dropping extra tied scores
    group_by(Event) %>% 
    slice_max(n = perevent, order_by = Score, with_ties = FALSE) %>% 
    mutate(
      teamscore = sum(Score),
      # names     = c(unique(NAME))
      )  %>%
    group_by(Event) %>% 
    summarise(
      Score   = first(teamscore),
      OnEvent = list((NAME))
      ) %>% ungroup()
}

# function to take output of select_top_scores and calculate total
team_total <- function(top_scores) {
  top_scores %>% 
    summarise(x=sum(Score)) %>% 
    pull()
}

```

This method assumes that scores earned at the USA Championships and Olympic
Trials are accurate predictions of scores earned at the actual Olympics.
These will be over-estimates of a potential team's Olympic performance if the 
domestic judging is less strict than the Olympic Committee judges panels.

My code also implicitly uses a 5-up, 3-count scoring system which is different
from the actual 3-up, 3-count rule during Olympic team finals.
While the US might choose the lineup in this way, they will also probably 
have to take into account the health of their athletes
(e.g., resting Brody's leg by not having him vault).

# Team USA's expected performance

## Event scores breakdown
```{r}
#| label: tbl-teamevents
#| tbl-cap: "Team USA expected event scores"
#| lst-label: lst-actualteam
#| lst-cap: team_top_scores() of actual team picks
team_names <- c("Brody Malone", "Fred Richard", "Asher Hong", "Paul Juda",
                "Stephen Nedoroscik")
actual_team <- team_top_scores(team_names)
actual_team_tot <- actual_team %>% 
  team_total()
actual_team
```

### Interpreting the team's average scores

The expected team score for this team ^[with domestic judging]
is `r actual_team_tot`.
For some context, the gold medal Japanese team at 2023 world championships
^[according to the [wikipedia page](https://en.m.wikipedia.org/wiki/2023_World_Artistic_Gymnastics_Championships)]
scored 255.594 and USA won bronze at the same competition with 252.428.

On paper this teams strengths relative to international competition seem to be
Floor, Pommel, and High Bar with scores of 
`r actual_team %>% filter(Event=="FX") %>% pull(Score)`,
`r actual_team %>% filter(Event=="PH") %>% pull(Score)`, and
`r actual_team %>% filter(Event=="HB") %>% pull(Score)`,
respectively.
However, the US is pretty infamous for not being able to pull together their 
full potential on Pommel.

This team also remains competitive on  Rings with an average score of 
`r actual_team %>% filter(Event=="SR") %>% pull(Score)`
and on P-Bars with a
`r actual_team %>% filter(Event=="PB") %>% pull(Score)`.
With China being so dominant on those two events they probably won't win any
event final medals there but having those two events as strengths will definitely help the team score against Japan and Great Britain.

It's kind of surprising that the main weakness on paper of this team is vault.
With Brody, Paul, and Asher vaulting at a total predicted score of 
`r actual_team %>% filter(Event=="VT") %>% pull(Score)`,
they fall behind both China's and Great Britain's scores as well as the 
previous US team's own score on the event from last year.
I heard that judging on Asher's Ri Se Gwang vault seemed overly harsh,
so it's possible that this team vault score gets better,
but I think he's the only one with a D-score higher than 5.6 on this event.

### Potential vs Expectations

There will also be more psychological strain on these athletes which comes with
the media attention and unfamiliar setting in Paris.
This could be a factor that could potentially affect this team more than other
alternative US teams.

I was hoping that Yul Muldauer could provide the same kind of leadership
he showed at the World Championships to a younger and less-internationally 
experienced team.
Unfortunately he didn't perform up to his usual level of execution and 
consistency in these past few competitions so I understand why he was left off.
These five guys (with maybe the exception of Paul?) have experience competing
at the World/Olympic level, but I think Yul is one of those guys who gets along
with everyone and feels comfortable in the spotlight.
I think Fred has a very similar enthusiasm and drive, but this is still only 
his second year of competing at the senior international level.

The first night of the Olympic Trials made me pretty nervous about the 
ability for US gymnasts to stay consistent with increased attention on them.
It was great to see how full the stadium in Minneapolis was for men's gymnastics,
but I also thought that some of the extra shakiness on Pommel and throughout the night might have been attributable to all the extra energy around.
This team has a lot of potential and I just hope that they don't get 
overwhelmed by the pressure of thinking that they need to live up to the 
high expectations of a domestic audience who only tune in every four years
with the anticipation that anything team USA touches at the Olympics could turn to gold.

### Comparing to the rest of the field

```{r}
#| warning: false
#| label: fig-eventdist_team
#| fig-cap: "Olympic team's scores relative to everyone else"
team_scores <- indv.averages_df %>% 
    filter(NAME %in% team_names) %>%
    select(-AA) %>% 
    pivot_longer(
      -NAME,
      names_to = "Event",
      values_to = "Score",
    )
eventsdist +
  geom_vline(
    data = team_scores, 
    aes(
      xintercept = Score, 
      color = Score,
      ),
      linetype = "dashed"
    ) +
  geom_text(
    data = team_scores, 
    aes(
      y = 0.4,
      x = Score,
      label = NAME,
      color = Score
      ),
    angle = 270,
    size = 2,
    nudge_x = 0.05,
    ) +
  scale_color_viridis_c(option = "turbo", direction = -1) +
  xlim(12, 15)
```

---

# Team Selection Analysis
```{r}
#| eval: false
#| label: sim_results

names <- indv.averages_df %>% select(NAME) %>% pull()

# all combinations of gymnasts choosing five at a time
teams_possible <- combinat::combn(names, 5, simplify = T) %>% 
  as_tibble() %>% 
  pivot_longer(
    cols = everything(),
    names_to = "team",
    values_to = "name") %>% 
  group_by(team) %>% 
  summarise(
    names = list(name)
  )

# function to apply team_total to list of names
team_top_total <- function(x) team_total(team_top_scores(unlist(x)))

# record top scores for every possible team combination
plan(multisession, workers = 10)
teamsim_df <- teams_possible %>% 
  mutate(
    total = as.numeric(furrr::future_map(names, team_top_total)),
    # names = map(names, toString)
  ) %>% 
  arrange(desc(total))

# save sim results to file
save(teamsim_df, file = "teamsim.RData")
```

```{r specialist_rule}
specialist_rule <- function(teams, individuals) {
  specialists <- individuals %>% 
    filter(is.na(AA)) %>% 
    select(NAME) %>% 
    pull()
  teams %>% 
    filter(
      str_count(.$names, paste(specialists, collapse='|')) < 2
    )
}
```


```{r}
#| label: tbl-topteams
#| tbl-cap: "Top ten scoring teams on average"
load("./teamsim.RData")
teamsim_df %>% 
  select(-team) %>% 
  filter(str_detect(.$names, "Patrick Hoopes", negate = T)) %>% 
  specialist_rule(individuals = indv.averages_df) %>% 
  head(n=10)
```

It's not surprising that Brody and Fred show up on all of these teams.
Asher is also pretty dominant in this ranking, which seems to be coming from his
higher performances on rings and p-bars compared to Donnell and being close
enough to Khoi on vault and floor.

The top team of all all-arounders just swaps out Stephen for Shane 
and scores more than a point lower.
Interesting that Yul has a better all-around average than Asher, 
and yet he wouldn't actually the better pick in an all all-arounder team.
A lot of these teams are just variations of the same teams except swapping 
Stephen and Patrick Hoopes.
Because Stephen had a higher four-score average and potential difficulty 
than Patrick, I drop Patrick from the list to get a sense of some more 
realistic alternate team selections.

This also highlights how close Shane came to being on the team (by just .75)! 
But it also how similar his contributions are to Paul's which sort of put him
in the same unfortunate position as Patrick Hoopes who's status as a pommel 
specialist means that he can only really be swapped out with Stephen.

### Importance of Pommel Horse Specialists

One thing I didn't expect before seeing these simulated teams
is that almost all of these top 20 teams have at least one pommel specialist
(before I dropped Patrick).

```{r}
#| echo: true
PH_actual_team <- actual_team %>% filter(Event=="PH") %>% 
  select(Score) %>% pull()
PH_second_best <- indv.averages_df %>% 
  filter(!is.na(AA)) %>% 
  select(NAME, PH) %>% 
  slice_max(order_by = PH, n = 3)
PH_second_best
```

```{r}
#| echo: false
PH_second_best_score <- PH_second_best %>% 
  summarise(x = sum(PH)) %>% 
  pull()
```

Without sending a pommel horse specialist to Paris, 
the best possible team score USA could put up on this event would be 
`r PH_second_best_score`.
So Stephen is contributing at least an extra 
`r PH_actual_team - PH_second_best_score` points to the team score
and likely more considering that he's making sure Fred's 13.3 PH average 
won't have to be counted. 

```{r}
#| echo: false
pull_score <- function(scores, name, event) {
  scores %>% 
    filter(NAME == name) %>% 
    select(event) %>% 
    pull()
}
```

If Khoi had hit his pommel routines consistently then he would have seemed
like a much more obvious replacement for Stephen because he has the potential
to put up a 15.0 like he showed in night 2 of US Championships.
He would also make the team's vault score a lot more competitive where his 
average score of 
`r pull_score(indv.averages_df, "Khoi Young", "VT")` is actually 
`r round(pull_score(indv.averages_df, "Khoi Young", "VT") - pull_score(indv.averages_df, "Asher Hong", "VT"), 1)*10`
tenths higher than Asher's and seems to be a lot more consistent.

# Other hypothetical compositions

### Picking only the top all-arounders
```{r}
#| label: tbl-altteam1events
#| tbl-cap: "Top 5 all-arounders team event scores"
# If they had taken top five all arounders
alt_team1_names <- indv.averages_df %>% 
  slice_head(n=5) %>% 
  select(NAME) %>% 
  pull()
alt_team1 <- team_top_scores(alt_team1_names)
alt_team1
```

This team has an expected score of `r team_total(alt_team1)` 
which is `r round(actual_team_tot - team_total(alt_team1), digits=1)` points 
behind the selected team's expected score,
so it makes sense they didn't go with this team.
Paul and Shane are both similar in that their main strengths are contributing 
solid performances on any event but neither have any standout routines with
enough difficulty to put them in event finals. 
It makes sense to have one gymnast fill this roll on a team, 
but I don't think it makes sense to leave behind world-class routines like 
Stephen's pommel horse or Asher's, Khoi's, or Donnell's vaults to gain
the additional safe scores from sending both Paul and Shane together.

It was really cool to see how hyped the Minneapolis crowd was to see 
shane compete in his hometown and he did a great job over the two nights there,
but ultimately Paul just put on a more consistent performance during 
US Championships. 
And that's what this type of role on team USA is all about;
just proving that you will put up a decent score every time you go up.

### My team picks
```{r}
#| label: tbl-altteam2events
#| tbl-cap: "My custom team's event breakdown"
# My own personal picks
alt_team2_names <- c("Fred Richard", "Brody Malone", "Yul Moldauer", "Donnell Whittenburg", "Khoi Young")
alt_team2 <- team_top_scores(alt_team2_names)
alt_team2
```

This team falls short by `r round(actual_team_tot - team_total(alt_team2), 1)` points
although I think they would have more opporunities to make event finals 
on Floor, P-Bars, and Vault.
Given Donnell's vs Asher's performances over Nationals and Trials,
it probably makes more sense to have Asher 
but I think Donnell has more international experience 
and I also just really wanted to see him get to represent USA at the Olympics
after working so hard for his entire career and not really getting that much recognition.

### Using Maximum Scores instead of Averages

```{r}
#| warning: false
indv.max_df <- results_df %>% 
  filter(NAME %in% ot1_df$NAME) %>%  # who qualified to olympic trials 
  group_by(NAME) %>% 
  summarise_at(vars(FX:AA), max, na.rm=TRUE) %>% 
  arrange(desc(AA))

team_max_total <- function(x) team_total(team_top_scores(unlist(x), scores = indv.max_df))
team_max_total(alt_team2)
```

```{r}
# record top scores for every possible team combination using individual max's
plan(multisession, workers = 10)
teamsim_df <- teamsim_df %>% 
  mutate(
    total_max = as.numeric(future_map(names, team_max_total)),
    # names = map(names, toString)
  ) %>% 
  arrange(desc(total_max))

save(teamsim_df, file = "teamsim.RData")

load("teamsim.RData")

teamsim_df %>% 
  select(-team) %>% 
  filter(str_detect(.$names, "Patrick Hoopes", negate = T)) %>% 
  specialist_rule(individuals = indv.averages_df) %>% 
  head(n=10)
```


# Small aside: runner-up team

I recently listened to Sam Oldham's podcast episode on the olympic team final 
with Hamish Carter and he highlighted the depth of team USA as opposed to 
GBR's current program by saying that the guys in US olympic trials who got left
off the final team could probably be their own olympic team.

I wanted to see who would be the next-best team after leaving out 
Brody, Frederick, Paul, Asher, and Stephen so I filtered them out of all the 
simulated team scores:
```{r}
load("./teamsim.RData")
teamsim_df %>% 
  filter(!str_detect(names, 'Malone|Richard|Juda|Hong|Nedoroscik')) %>% 
  head(5) %>% 
  kableExtra::kable()
```

These averages are just below Team GBR's Olympic team final performance in Paris
but they have a very competitive vault and p-bar line-up.

```{r}
runnerup_team <- c("Shane Wiskus","Donnell Whittenburg", "Cameron Bock", "Khoi Young", "Curran Phillips")
team_top_scores(runnerup_team) %>% kableExtra::kable()
```

I can also compare these gymnasts' top individual apparatus scores to the 
Olympic event finals to see who would have been in contention for medals.

```{r}
olympic_results_df <- read_csv("~/Personal/MAG-stats/results/olympics-2024.csv") %>% 
  filter(!ROUND %in% c("All-Around Final Results"))

runnerup_v_olympics <- results_df %>% 
  filter(NAME %in% runnerup_team) %>% 
  group_by(NAME) %>% 
  summarise(
    across(FX:HB, ~ max(., na.rm = T))
  ) %>% 
  bind_rows(olympic_results_df)

runnerup_v_olympics %>% 
  select(NAME, FX) %>% 
  arrange(desc(FX))

runnerup_v_olympics %>% 
  select(NAME, PH) %>% 
  arrange(desc(PH))

runnerup_v_olympics %>% 
  select(NAME, SR) %>% 
  arrange(desc(SR))

runnerup_v_olympics %>% 
  select(NAME, VT) %>% 
  arrange(desc(VT))

runnerup_v_olympics %>% 
  select(NAME, PB) %>% 
  arrange(desc(PB))

runnerup_v_olympics %>% 
  select(NAME, PB) %>% 
  arrange(desc(PB))
```

From this, it seems like Khoi on Vault and Curran on P-bars would have been 
the main two individual medal contenders.
It's too bad that one of the best parallel bar workers in the world right now
was left out of the olympics and probably won't get an opportunity to earn any 
world championships medals either.
I think Khoi still has a lot of time to keep getting more consistent and after 
winning multiple medals at the 2023 world championships I think he will get
more opportunities to grow as an international-level gymnast.
He was probably under a lot of pressure to up his game even more during the 
team selection processs after such a successful international and collegieate 
season in the year prior to Paris.
But I think that he will probably be an important part of the future of the US
team on the road to LA 2028.

Unfortunately for Donnell, Shane, and Cameron, I think Paris was probably their 
last shot at olympic teams. However, all three have already had long and 
internationally successful gymnastics careers.