---
title: "Olympics Pommel Medal Predictions"
author: "Dante Yasui"
date: "2024/07/21"
format:
  html:
    df-print: kable
    toc: true
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

olympians <- read_csv("./olympic_teams.csv") %>%
  select(NAME) %>% 
  pull()

results_df <- read_csv("./2024-aggregate-results.csv") %>% 
  filter( 
    !(MEET %in% c("world_champs")),
    NAME %in% olympians
    )

indv.averages_df <- results_df %>% 
  group_by(NAME, TEAM) %>% 
  summarise(
    PH = mean(PH, na.rm=TRUE),
    PH_D = max(PH_D, na.rm=TRUE)
    )%>% 
  mutate(PH_D = na_if(PH_D, -Inf)) %>% 
  arrange(desc(PH)) 
```

```{r}
indv.averages_df %>% 
  head(n = 8)
```

## Selected Routines

### Nariman Kurbanov
{{< video https://www.youtube.com/watch?v=FOJEHyE1hsU >}}

<<<<<<< HEAD
```{r}
results_df %>% 
  filter(NAME == "Nariman Kurbanov") %>% 
  select(MEET, ROUND, PH, PH_D)
```


### Stephen Nedoroscik
{{< video https://www.youtube.com/watch?v=qHYwwLzQ9Y0 >}}

```{r}
results_df %>% 
  filter(NAME == "Stephen Nedoroscik") %>% 
  select(MEET, ROUND, PH, PH_D)
```
=======
### Stephen Nederoscik
{{< video https://www.youtube.com/watch?v=qHYwwLzQ9Y0 >}}

### Lee Chih Kai
{{< video https://www.youtube.com/watch?v=uWOYqXL6_rY >}}
>>>>>>> d35f0d043a5a73837da3504c7c38c1acdb5a9236

### Max Whitlock
{{< video https://www.youtube.com/watch?v=qY_bIeFrj3s >}}

<<<<<<< HEAD
```{r}
results_df %>% 
  filter(NAME == "Max Whitlock") %>% 
  select(MEET, ROUND, PH, PH_D) %>% 
  filter(!is.na(PH))
```

### Ahmad Abu Al Soud
{{< video https://www.youtube.com/watch?v=yypfmxI2e-Y >}}

```{r}
results_df %>% 
  filter(NAME == "Ahmad Abu Al Soud") %>% 
  select(MEET, ROUND, PH, PH_D) %>% 
  filter(!is.na(PH))
```

### Rhys McClenaghan
{{< video https://www.youtube.com/watch?v=CaPLLtpWTVQ >}}

```{r}
results_df %>% 
  filter(NAME == "Rhys McClenaghan") %>% 
  select(MEET, ROUND, PH, PH_D) 
```
=======
### Ahmad Abu Al Soud
{{< video https://www.youtube.com/watch?v=yypfmxI2e-Y >}}
>>>>>>> d35f0d043a5a73837da3504c7c38c1acdb5a9236
